---
title: "Building a manuscript table end to end"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building a manuscript table end to end}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(hvtiRtables)
```

This walks one complete JTCVS submission table, from a data frame to a
checked `.docx`. The CORR house-style path is shorter and covered in
`?hv_man_table`.

## 1. Summarize the data

`hv_tbl_summary()` takes a grouped, ordered variable list and type
buckets, the way the SAS `%summarytable` macro does. Every variable in
`groups` appears in exactly one of `continuous`, `binary`, or
`categorical`.

```{r}
tbl <- hv_tbl_summary(
  gtsummary::trial,
  by = "trt",
  groups = list(
    Demography = c("age", "marker"),
    Disease = c("stage", "grade")
  ),
  continuous = c("age", "marker"),
  categorical = c("stage", "grade"),
  compare = "pvalue"
)
```

The result carries two attributes the renderer wants: the
percentile-aware sub-header text, and a ready-made `trailing` argument
for the comparison column.

```{r}
attr(tbl, "hv_stat_label")
attr(tbl, "hv_trailing")
```

## 2. Render the JTCVS shape

```{r}
ft <- hv_man_table_jtcvs(
  tbl,
  groups = c(stat_1 = "Drug A (n=98)", stat_2 = "Drug B (n=102)"),
  stat_label = attr(tbl, "hv_stat_label"),
  trailing = attr(tbl, "hv_trailing")
)
```

`groups` names are the `stat_<k>` columns `gtsummary` built, one per
level of `trt`. The labels are yours, including each arm's N.

## 3. Build the test footnotes

`hv_test_footnotes_jtcvs()` reads which test produced each p-value and
returns footnotes in the shape the saver expects, with the body-row
indices already computed.

```{r}
notes <- hv_test_footnotes_jtcvs(tbl)
length(notes)
notes[[1]]$text
```

## 4. Save and check

```{r}
out <- tempfile(fileext = ".docx")
hv_man_table_save_jtcvs(
  ft, out,
  caption = "Table 1. Baseline Characteristics",
  footnotes = notes,
  abbreviations = c(SMD = "standardized mean difference")
)
```

Neither saver checks its own output, so run `hv_check_docx()` on the
path. Zero rows means clean.

```{r}
hv_check_docx(out)
```

## What the package refuses

Every public function rejects out-of-spec input with a message naming
the fix, rather than letting it reach `gtsummary` or `flextable`.

```{r, error = TRUE}
hv_man_table_jtcvs(
  gtsummary::tbl_summary(gtsummary::trial, by = "trt", include = "age"),
  groups = c(stat_1 = "A", stat_2 = "B")
)
```

```{r, error = TRUE}
hv_tbl_summary(
  gtsummary::trial, groups = list(D = "age"),
  continuous = "age", binary = "age"
)
```
