---
title: "Porting a %summarytable program to R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Porting a %summarytable program to R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(hvtiRtables)
```

This is written for someone who knows `%summarytable` and the `tp.dc.*`
templates, and does not want to learn `gtsummary` beyond what the port
requires. It ends at the summary object; rendering it is covered in
`vignette("hvtiRtables")`.

## One macro call becomes three

`%summarytable` computes the table *and* writes the document. Its
`RTFFILE=`, `TBLTITLE=`, `ADDFN=`, `STYLE=` and `PAGE=` parameters sit in
the same call as `CON1=` and `CLASS=`. This package splits that work, so
the parameters you know are distributed across separate calls:

| Stage | Function | Macro parameters that land here |
|---|---|---|
| Compute stats and tests | `hv_tbl_summary()` | `DATA=`, `CLASS=`, `CON*=`, `CAT*=`, `ORD1=`, `LIST=`, `PP=`, `PVALUES=`, `ASD=`, `TOTALCOL=` |
| Shape the table | `hv_man_table()`, `hv_man_table_jtcvs()` | *(none - house style is fixed)* |
| Write the file | `hv_man_table_save()`, `hv_man_table_save_jtcvs()` | `RTFFILE=`, `PDFFILE=`, `XMLFILE=`, `TBLTITLE=`, `ADDFN=`, `PRINTFN=` |
| Verify the file | `hv_check_docx()` | *(no analogue)* |

If you pass a macro parameter to the wrong stage, the error says which
function takes it:

```{r, error = TRUE}
hv_tbl_summary(
  gtsummary::trial, by = "trt",
  groups = list(Demography = "age"), continuous = "age",
  tbltitle = "Table 1"
)
```

## Defaults that differ

Five places where a faithful-looking port produces different numbers.

| Macro | Default there | Default here | Fix |
|---|---|---|---|
| `TOTALCOL=` | `1` - Overall column included | `overall = FALSE` | `overall = TRUE` |
| `CON1=` | mean +/- SD, one-way ANOVA | median (percentiles), Kruskal-Wallis | none - see below |
| `PP=` | `16 84` nominal; house templates use both `15 85` and `16 84` | `percentiles = c(15, 85)` | `percentiles = c(16, 84)` |
| `CUTEXACT=` | Fisher when >=50% of cells have expected < 5 | Fisher when *any* expected < 5 | none - not tunable |
| `ORD1=` | `n (%)`, Kruskal-Wallis | folds into `categorical`, chi-square | none |

`overall = FALSE` is the one deliberate departure from a macro default.
The renderers take a `groups` vector naming each `stat_<k>` column, so a
column appearing by default would break every existing call.

## A ported call

The `%summarytable` call from `tp.dc.stddiff.summarytable.sas`:

```sas
%summarytable(data=built,
              class=grp_res,
              con3=&nongau.,
              cat1=&binary.,
              cat2=&catg.,
              pp=15 85,
              totalcol=0,
              pvalues=1,
              list=
    /* Demography */
       female AGE BSA race_gp
    /* Symptoms */
       surgstat nyha_pr
);
```

and its equivalent. Note that `pp=` and `list=` map straight across,
`totalcol=0` is the default here so it disappears, and `pvalues=1` becomes
`compare = "pvalue"`:

```{r eval = FALSE}
tbl <- hv_tbl_summary(
  built,
  by = "grp_res",
  groups = list(
    Demography = c("female", "AGE", "BSA", "race_gp"),
    Symptoms   = c("surgstat", "nyha_pr")
  ),
  continuous  = c("AGE", "BSA"),
  binary      = c("female", "surgstat"),
  categorical = c("race_gp", "nyha_pr"),
  compare     = "pvalue",
  percentiles = c(15, 85)
)
```

## Parameter map

| `%summarytable` | Here | Notes |
|---|---|---|
| `DATA=` | `data` | |
| `CLASS=` | `by` | |
| `LIST=` | `groups` | a named list; the section comments become the names |
| `CON1=` | `continuous` | **statistic changes** - mean +/- SD becomes median (percentiles) |
| `CON2=` | `continuous` | **statistic changes** - median (min, max) becomes median (percentiles) |
| `CON3=` | `continuous` | direct equivalent; this is the behavior `continuous` reproduces |
| `CAT1=` | `binary` | |
| `CAT2=` | `categorical` | |
| `ORD1=` | `categorical` | **test changes** - Kruskal-Wallis becomes chi-square |
| `PP=` | `percentiles` | |
| `PVALUES=` | `compare = "pvalue"` | |
| `ASD=` | `compare = "smd"` / `"both"` | |
| `TOTALCOL=` | `overall` | default differs - see above |
| `NCOL=` | automatic | per-group Ns always shown; `NCOL=3`'s overall N needs `overall = TRUE` |
| `TBLTITLE=` | `caption` | **JTCVS only** - `hv_man_table_save()` writes no caption |
| `ADDFN=` | `footnotes` | on the save function |
| `PRINTFN=` | `footnotes = hv_man_footnotes()` | **CORR only.** The house block is CORR-shaped and cannot be passed to `hv_man_table_save_jtcvs()`, which needs `list(row =, col =, text =)` entries; build those, or use `hv_test_footnotes_jtcvs()`. It also hard-codes the 15th/85th percentiles, so change it if you change `percentiles` |
| `RTFFILE=`, `PDFFILE=`, `XMLFILE=` | `file` | output is always `.docx` |

The save stage is a sibling pair, and the two do not take the same
arguments: only `hv_man_table_save_jtcvs()` has `caption`. For a flat-header
CORR table the title goes in the document, not the call.

Not supported: `WEIGHT=`, `PROPENMT=`, `MWOUTCOMES=`, `SUBSET=` (subset the
data frame first), `SORTBY=` (order comes from `groups`), `COLPCT=0` (row
percentages), `MISSCOL=`, `ADHOC=`, `ODDSRATIOS=`, `CUTEXACT=`, `STYLE=`,
`PAGE=`, `CWIDTH1-3=`, `PAPERSIZE=`.

## Why my p-value moved

Three causes, in order of how often they explain it.

1. **A `CON1=` variable.** The macro ran one-way ANOVA on it. Every
   continuous variable here is tested non-parametrically. This is also why
   no footnote letter reads `a=ANOVA`.
2. **An `ORD1=` variable.** The macro ran Kruskal-Wallis; folded into
   `categorical`, it gets chi-square. No trend test is run.
3. **A sparse categorical variable.** Both switch to Fisher's exact, at
   different thresholds - `CUTEXACT=50` needs half the cells to have
   expected counts below 5, `gtsummary` needs one.

## Percentiles and QNTLDEF

SAS's house `QNTLDEF=5` is not R's default quantile definition. It
corresponds to `type = 2`:

```{r}
stats::quantile(c(1, 2, 3, 4), 0.25, type = 2, names = FALSE)  # SAS house
stats::quantile(c(1, 2, 3, 4), 0.25, type = 7, names = FALSE)  # R default
```

Medians agree at every n; Q1, Q3 and `pNN` diverge silently. `gtsummary`'s
`{pXX}` statistics already use `type = 2`, so `hv_tbl_summary()` matches
SAS without intervention. **Any `stats::quantile()` call you write by hand
in ported code does not** - pin `type = 2` explicitly.

The percentiles actually used are recorded on the table object as
`attr(tbl, "hv_stat_label")`, but nothing consumes that attribute
automatically. `hv_man_table_jtcvs()`'s `stat_label` argument defaults to
a literal mean-and-SD string, so a `percentiles = c(16, 84)` table
rendered without an explicit `stat_label =` claims mean/SD over median
columns. You MUST pass it yourself:

```{r}
tbl <- hv_tbl_summary(
  gtsummary::trial, by = "trt",
  groups = list(Demography = c("age", "marker")),
  continuous = c("age", "marker"),
  percentiles = c(16, 84)
)
attr(tbl, "hv_stat_label")
```

```{r eval = FALSE}
ft <- hv_man_table_jtcvs(
  tbl,
  groups = c(stat_1 = "Drug A", stat_2 = "Drug B"),
  stat_label = attr(tbl, "hv_stat_label")
)
```

## Reading the SAS data in

`haven::read_sas()` returns `haven_labelled` columns over doubles, because
every SAS numeric is an 8-byte float. A 0/1 flag therefore arrives looking
like a continuous variable, and type guessers classify it as one.

This is why `binary` and `categorical` are stated rather than inferred:
name every variable in exactly one of `continuous`, `binary` or
`categorical`, and a misclassification is an error rather than a wrong
table.

```{r, error = TRUE}
hv_tbl_summary(
  gtsummary::trial, groups = list(D = "age"),
  continuous = "age", binary = "age"
)
```
